Rough calculations

### 2015-11-02 notes/diffusion-local-times.Rmd
.

Let's compare the case where the cline is set up instantly or not:
```{r compare_rightness, cache=TRUE,dependson=c("fwds_back_pde","solve_pde_setup","setup_fns")}
layout(t(1:2))
fb.p <- clineplot(rev.soln,main="forwards-backwards")
st.p <- clineplot(pde.soln,main="stationary cline")
layout(1)
plot(fb.p,st.p,asp=1,pch=20,cex=0.5,xlab="forwards-backwards",ylab="stationary",col=rainbow(ncol(fb.p))[col(fb.p)])
abline(0,1,lwd=3)
```

>>>>

Let's compare the case where the cline is set up instantly or not:
```{r compare_rightness, cache=TRUE,dependson=c("fwds_back_pde","solve_pde_setup")}
layout(t(1:2))
fb.p <- clineplot(rev.soln,main="forwards-backwards")
st.p <- clineplot(pde.soln,main="stationary cline")
layout(1)
plot(fb.p,st.p,asp=1,pch=20,cex=0.5,xlab="forwards-backwards",ylab="stationary",col=rainbow(ncol(fb.p))[col(fb.p)])
abline(0,1,lwd=3)
```

### 2015-11-02 notes/pde-clines.Rmd
.

We know that 
$$\begin{aligned}
q(x,A,0,r) = 1  \\
q(x,B,0,r) = 0  \\
\end{aligned}$$
and that $\partial_t q = G q$, i.e.,
(writing $q(x,A,t,r) = q_A(x,t)$ for the moment):
$$\begin{aligned}
    \partial_t q_A(x,t)
        &= 
        \grad \log(p(x)) \cdot \grad q_A(x,t) + \frac{\sigma^2}{2} \Delta q_A(x,t) + r (1-p(x)) (q_B(x,t)-q_A(x,t)) \\
    \partial_t q_B(x,t)
        &= 
        \grad \log(1-p(x)) \cdot \grad q_B(x,t) + \frac{\sigma^2}{2} \Delta q_B(x,t) + r p(x) (q_A(x,t)-q_B(x,t)) .
\end{aligned}$$

@@

>>>>

We know that 
$$\begin{aligned}
q(x,A,0,r) &= 1  \\
q(x,B,0,r) &= 0  \\
\end{aligned}$$
and that $\partial_t q = G q$, i.e.,
(writing $q(x,A,t,r) = q_A(x,t)$ for the moment):
$$\begin{aligned}
    \partial_t q_A(x,t)
        &= 
        \grad \log(p(x)) \cdot \grad q_A(x,t) + \frac{\sigma^2}{2} \Delta q_A(x,t) + r (1-p(x)) (q_B(x,t)-q_A(x,t)) \\
    \partial_t q_B(x,t)
        &= 
        \grad \log(1-p(x)) \cdot \grad q_B(x,t) + \frac{\sigma^2}{2} \Delta q_B(x,t) + r p(x) (q_A(x,t)-q_B(x,t)) .
\end{aligned}$$

@@

Haplotype lengths
=================

There are various haplotype-related statistics we could compute:

1.  Distance to closest recombination in a heterozygote

Distance to closest recombination in a heterozygote
---------------------------------------------------

The first haplotype-related statistic
is the distance along the chromosome from an $A$ allele until the closest recombination
that occurred in an $AB$ heterozygote.
Let $H$ be this distance;
then, 
$$\begin{aligned}
    u(x,\tau,r)=\P^(x,\tau)\{H>r\}
\end{aligned}$$
solves the following PDE:
$$\begin{aligned}
    u(x,0,r) &= 1  \\
    \partial_t u(x,t,r)
        &= 
        \grad \log(p(x)) \cdot \grad u(x,t,r) + \frac{\sigma^2}{2} \Delta u(x,t,r) - r (1-p(x)) u(x,t,r) .
\end{aligned}$$

### 2015-11-02 notes/rough-calculations.Rmd
.

1.  What is the width of the cline at a linked, neutral locus?
    How long does it take to flatten out?
    How far from the cline do you have to go to find mostly one type?

2.  How long are the haplotypes around the rare $B$ alleles on the $A$ side?
    This should be constant in time.

3.  How long, and how common, are chunks of $B$ haplotype *not* linked to $B$ alleles on the $A$ side?
    These will accumulate as time progresses.

4.  How does neighborhood size affect the results?
    Does the mean cline, averaged across low-population-density simulations,
    match the cline obtained by a high population density, i.e., the PDE?

5.  What about the variance??

6.  For a pair of sites, what is the correlation in ancestral state at those loci,
    given the state at the selected locus
    (and, also, averaged over choice of selected locus).

@@

@@

a.  Maximum cline width $\sigma \sqrt{\tau}$, even at unlinked loci.
b.  The selected cline takes time $s$ to spread out to its full width.
c.  Cline behaves like the selected cline for $r < 1/\tau$ (or, for $r < \sqrt{\frac{2}{s\tau}}$ if $\tau>1/s$).
d.  Cline is flattening for $x<\sigma\sqrt{\tau}$ and $r > \sqrt{\frac{2}{s\tau}}$.

*Rationale:*

@@

b.  The selected cline spreads out (in forwards time!) with diffusion,
    and so will reach maximum width $\sigma/\sqrt{s}$ in time $s$.
    Let $q(t,x)$ be the solution to
    $$\begin{aligned}
    \dot q = \frac{1}{2} q'' + q (1-q) (2q-1) .
    \end{aligned}$$
    If $u=at$ and $y=bx$ and let $p(t,x) = q(at,bx)$ then
    $$\begin{aligned}
    \dot p(t,x)
        &= \partial_t q(at,bx) \\
        &= a \dot q(at,bx) \\
        &= a \left\{ \frac{1}{2} q''(at,bx) + q (1-q) (2q-1) \right\} \\
        &= a \left\{ \frac{1}{2b^2} \partial^2_x q(at,bx) + q (1-q) (2q-1) \right\} \\
        &= \frac{a}{2b^2} p''(t,x) + a p (1-p) (2p-1)  .
    \end{aligned}$$
    If we choose $a=s$ and $b=\sigma/\sqrt{s}$ then $a/b^2=\sigma^2$, so
    $$\begin{aligned}
    \dot p = \frac{\sigma^2}{2} p'' + s p (1-p) (2p-1) .
    \end{aligned}$$
    We know that the unitless cline $q$ will spread out in time 1;
    so our cline will spread out in time $s$.

@@

c.  Look at space-recombination frequency images at a bunch of times
    to check that flattening occurs around $r=\sqrt{2/s\tau}$.

>>>>

1.  **How a neutral cline flattens:**  
    What is the width of the cline at a linked, neutral locus?
    How long does it take to flatten out?
    How far from the cline do you have to go to find mostly one type?

2.  **How do haplotypes look around the selected locus in the cline:**
    How long are the haplotypes around the rare $B$ alleles on the $A$ side?
    This should be constant in time.

3.  **How does introgression happen:**
    How long, and how common, are chunks of $B$ haplotype *not* linked to $B$ alleles on the $A$ side?
    These will accumulate as time progresses.

4.  **What can we learn from linkage?**
    For a pair of sites, what is the correlation in ancestral state at those loci,
    given the state at the selected locus
    (and, also, averaged over choice of selected locus).

@@

5.  How does neighborhood size affect the results?
    Does the mean cline, averaged across low-population-density simulations,
    match the cline obtained by a high population density, i.e., the PDE?

6.  What about the variance??

@@

a.  Maximum cline width $\sigma \sqrt{\tau}$, even at unlinked loci.
b.  The selected cline takes time $1/s$ to spread out to its full width.
c.  Cline behaves like the selected cline for $r < 1/\tau$, and begin to flatten for $r>1/\tau$.
d.  For large distances and times (but with $x<\sigma\sqrt{\tau}$),
    the cline may stay steep out to
    $r = \sqrt{\frac{2s}{\tau}}$ rather than $r=1/\tau$,
    but this seems hard to actually observe.

**Rationale:**

@@

b.  The selected cline spreads out (in forwards time!) with diffusion,
    and so will reach maximum width $\sigma/\sqrt{s}$ in time $1/s$.
    Let $q(t,x)$ be the solution to
    $$\begin{aligned}
    \dot q = \frac{1}{2} q'' + q (1-q) (2q-1) .
    \end{aligned}$$
    If $u=at$ and $y=bx$ and let $p(t,x) = q(at,bx)$ then
    $$\begin{aligned}
    \dot p(t,x)
        &= \partial_t q(at,bx) \\
        &= a \dot q(at,bx) \\
        &= a \left\{ \frac{1}{2} q''(at,bx) + q (1-q) (2q-1) \right\} \\
        &= a \left\{ \frac{1}{2b^2} \partial^2_x q(at,bx) + q (1-q) (2q-1) \right\} \\
        &= \frac{a}{2b^2} p''(t,x) + a p (1-p) (2p-1)  .
    \end{aligned}$$
    If we choose $a=s$ and $b=\sigma/\sqrt{s}$ then $a/b^2=\sigma^2$, so
    $$\begin{aligned}
    \dot p = \frac{\sigma^2}{2} p'' + s p (1-p) (2p-1) .
    \end{aligned}$$
    We know that the unitless cline $q$ will spread out in time 1;
    so our cline will spread out in time $1/s$.

@@

c.  To check that flattening occurs around $r=1/\tau$ (i.e., at $\tau=1/r$)
    we'll look at when a neutral allele linked to the $A$ allele
    near the center starts to get close to 1/2.

    (*Technical note:* the system of ODE is [stiff](https://en.wikipedia.org/wiki/Stiff_equation), 
    as diffusive systems [usually are](http://scicomp.stackexchange.com/questions/891/the-definition-of-stiff-ode-system),
    particularly for small $r$ at the times when each set of probabilities becomes steep;
    setting $s=0.1$ this is a problem, but not with smaller (less steep) $s$.)

    ```{r space_recomb, cache=TRUE}
    s <- 0.01
    rvals <- c(0,exp(seq(log(.0025),log(0.05),length.out=8)))
    tt <- seq(0,1000,length.out=100)
    thegrid <- setup.grid.1D(x.up=-50, x.down=50, N = 2000)
    fwds.soln <- forwards_pde(s=s,times=seq(0,1.2*max(tt),length.out=1000),grid=thegrid)
    r.solns <- lapply(rvals, function(r) {
            forwards_backwards_pde(s=s,times=tt,grid=thegrid,r=r,log.p=TRUE,fwds.soln=fwds.soln)
        } )
    linked.clines  <- lapply(r.solns, cline_from_soln, grid=thegrid )
    xr <- lapply( seq_along(tt), function (k) { do.call( cbind, lapply( linked.clines, function (x) x[k,-1] ) ) } )
    ```
    The vertical lines are at $1/r$ (only on the righthand plots, due to the limitations of `filled.contour`):
    ```{r plot_sr}
    for (k in seq_along(rvals)) {
        image( r.solns[[k]], legend=TRUE, zlim=c(0,1), yaxt='n', ylab='position', main=sprintf("r=%f, s=%f",rvals[k],s=s) )
        axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
        abline(v=1/rvals[k],lwd=2,lty=3)
    }
    ```

d.  To look for the $r=1/\sqrt{\tau}$ behaviour, we could look at positions outside the selected cline 
    ($x$ larger than $C/\sqrt{2s}$, for C=4, say) 
    and times large enough that unlinked clines have started to flatten out there 
    and that the selected cline has got to full width
    ($\tau$ larger than $C^2/s$).
    But: if $\tau=C^2/s$ then the critical value is $r=(s/C)^2$,
    which at $s=0.01$ is $r=6.25\times10^{-6}$
    **Moral:** the narrower cline due to linkage may be only visible only very nearby to the selected site.
    *To-do:* figure out parameter values where this behaviour becomes visible: need to be much further away?

<!--
    ```{r long_time,cache=TRUE}
    s <- 0.01
    r <- 0.0001
    tt <- seq(0,30000,length.out=500)
    thegrid <- setup.grid.1D(x.up=-200, x.down=200, N = 2000)
    fwds.soln <- forwards_pde(s=s,times=seq(0,1.2*max(tt),length.out=1000),grid=thegrid)
    rev.soln <- forwards_backwards_pde(s=s,times=tt,grid=thegrid,r=r,log.p=TRUE,fwds.soln=fwds.soln)
    linked.cline  <- cline_from_soln( rev.soln, grid=thegrid )
    ```
-->

How do haplotypes look around the selected locus in the cline
=============================================================

2.  How long are the haplotypes around the rare $B$ alleles on the $A$ side?

a.  The haplotype of a $B$ allele at distance $x$ into the $A$ side should be of length about $\sigma/(sx)$.

**Rationale:**

a.  This is (probably) the Laplace transform of the time since the $B$ allele
    last had an ancestor on the $B$ side,
    i.e., for a $B$ lineage to make it back to the $B$ side.
    Since lineages on the wrong side move at speed $\sigma/s$,
    if the allele is at position $x$, 
    this will take about $sx/\sigma$;
    and so the haplotype should be about $\sigma/(sx)$.
